### 2013-08-22 doc/dissertation/document/backmatter/mcsa_spn_estimator_analysis.tex
moving spn estimator analysis to an appendix

\chapter{Monte Carlo Synthetic Acceleration\\ Methods for the $SP_N$ Equations}
\label{ch:spn_equations}

In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
equations, closely following the work of Evans
\cite{evans_simplified_2013}, in order to gain full understanding of
the underlying system and its potential behavior in a Monte Carlo
context. From the $P_N$ equations, we apply a set of approximations to
yield the $SP_N$ equations for fixed source and criticality
problems. Using the fully-formed linear operator for the transport
problem, we explore solutions to the $SP_N$ equations with Monte Carlo
Synthetic Acceleration using a challenging light water reactor fuel
assembly criticality calculation as the driving problem. Several
difficulties arise when applying MCSA to these problems that were not
observed when investigating the simple transport system in the
previous chapter.

For Jacobi-based preconditioners, convergence with MCSA is difficult
if not impossible for ill-conditioned systems with this behavior
demonstrated using a simple neutron diffusion problem. In order to
effectively solve the $SP_N$ equations for the fuel assembly problem,
a suite of preconditioners along with a relaxation scheme is developed
and studied within the context of MCSA. Using these preconditioners,
several additional issues are observed and alleviated to a certain
extent by applying the reduced domain approximation. MCSA solutions
for the fuel assembly problem are then verified by comparing the
solutions against production Krylov solvers for problems with varying
numbers of energy groups. Finally performance is analyzed through
comparison with those same Krylov solvers in terms of both iterative
performance and CPU timing using the same set of problems utilized for
the verification.

%%---------------------------------------------------------------------------%%
\section{The Neutron Transport Equation}
\label{sec:transport_eq}
As a starting point for the $SP_N$ equations we define the
time-independent neutron transport equation
\cite{lewis_computational_1993}:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
  q(\vec{r},\hat{\Omega},E)\:,
  \label{eq:general_transport}
\end{multline}
with the variables defined as:
\begin{itemize}
\item $\vec{r}$ - neutron spatial position
\item $\hat{\Omega}$ - neutron streaming direction with radial
  component $\mu$ and azimuthal component $\omega$
\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
  scattering
\item $E$ - neutron energy
\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
\item $\sigma(\vec{r},E)$ - total interaction cross section
\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
  of scattering from direction $\hat{\Omega}'$ into an angular domain
  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
  $E'$ to an energy domain $dE'$ about energy $E$
\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
\end{itemize}
For this work, it is sufficient to formulate
Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
  \frac{q(x,E)}{4 \pi}\:,
  \label{eq:cart_1d_transport}
\end{multline}
where the angular component of the solution is no longer dependent on
the azimuthal direction of travel and an isotropic source of neutrons
is assumed. In addition, for fission systems the eigenvalue form of
the transport equation is:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
  \label{eq:eigenvalue_transport}
\end{multline}
with the additional variables defined as
\begin{itemize}
\item $k$ - multiplication factor
\item $\chi(E)$ - fission neutron energy spectrum
\item $\nu$ - average number of neutrons per fission
\item $\sigma_f(r,E')$ - fission cross section\:.
\end{itemize}
In 1-dimensional Cartesian geometry,
Eq~(\ref{eq:eigenvalue_transport}) becomes:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
  dE' + \frac{q(x,E)}{4 \pi}\:.
  \label{eq:cart_1d_eigenvalue}
\end{multline}

%%---------------------------------------------------------------------------%%
\section{Derivation of the Monoenergetic $SP_N$ Equations}
\label{sec:spn_equations}
The $P_N$ equations as derived in \cite{lewis_computational_1993}
give $N+1$ coupled first-order equations capturing the spatial and
angular-dependence of the solution:
\begin{equation}
   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
  \label{eq:final_pn_equations}
\end{equation}
where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
yields a set of $N+1$ equations for $N+2$ flux moments with the
highest order moment set to zero for closure. In multiple dimensions,
the equation set becomes large and coupled not only through angular
moments but also through the spatial variables. As a simpler
alternative to multidimensional $P_N$ solutions, Gelbard recognized in
1960 that the planar $P_N$ equations could be simplified and applied
an ad-hoc method to extend them to multiple dimensions, yielding the
$SP_N$ equations. These equations are not only fewer in number, but
also take on a diffusion-like form while maintaining the angular
character of the flux, making them amenable to solutions with modern
diffusion methods.

First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
equations by solving for the $n^{th}$ Legendre flux moment in the
odd-order equations:
\begin{equation}
  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
  \label{eq:odd_moments}
\end{equation}
for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
reduced group of equations for the even moments:
\begin{multline}
  -\frac{\partial}{\partial x}
  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
  0,2,4,\cdots,N\:.
  \label{eq:reduced_pn}
\end{multline}
To extend these equations to multiple dimensions, Gelbard simply
replaced the planar spatial derivatives in the reduced set of
equations with general multidimensional gradient operators:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
  0,2,4,\cdots,N\:,
  \label{eq:spn_equations}
\end{multline}
yielding a multidimensional set of $(N+1)/2$ angular coupled equations
defined as the $SP_N$ equations. As with the $P_N$ equations, we
provide closure to this set of equations with $\phi_{N+1} = 0$. As a
concrete example, we will consider the $SP_7$ equations:
\begin{subequations}
  \begin{gather}
    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
    \Sigma_0 \phi_0 = q \\ 
    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
    \Sigma_2 \phi_2 = 0\\
    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
    \Sigma_6 \phi_6 = 0 \:.
  \end{gather}
  \label{eq:sp7_equations}
\end{subequations}
To further modify these equations, we can use a change of variables to
create a new group of equations such that the gradients are operating
on a single vector:
\begin{subequations}
  \begin{gather}
    u_1 = \phi_0 + 2\phi_2 \\
    u_2 = 3\phi_2 + 4\phi_4 \\
    u_3 = 5\phi_4 + 6\phi_6 \\
    u_4 = 7\phi_6 \:.
  \end{gather}
  \label{eq:spn7_subs}
\end{subequations}
When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
\begin{subequations}
  \begin{gather}
    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
    = -q \\
    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
    \frac{1}{7}u_4 \Bigg] = 0 \:.
  \end{gather}
  \label{eq:spn7_subs_equations}
\end{subequations}
If we rearrange Eq~(\ref{eq:spn7_subs_equations}) such that only one
divergence operation is present in each equation, we can formulate
this as a matrix system of 4 equations in the case of the $SP_7$
approximation:
\begin{equation}
  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
  \label{eq:spn_matrix}
\end{equation}
with $\mathbf{u}$ the vector of solution variables:
\begin{equation}
  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
  \label{eq:spn7_solution_vector}
\end{equation}
$\mathbf{D}$ the vector of effective diffusion coefficients:
\begin{equation}
  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
  \label{eq:spn7_diffusion_coeffs}
\end{equation}
$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
source has now been distributed through the system:
\begin{equation}
  \mathbf{q} = (
  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
  \label{eq:spn7_source_vector}
\end{equation}
and $\mathbf{A}$ a matrix of angular scattering terms:
% NOTE: I copied the following matrix directly out of Tom's tech note
% on the SPn equations.
\begin{equation}
  \mathbf{A} = 
  {\tiny \begin{bmatrix}
    (\Sigma_0) &
    (-\frac{2}{3}\Sigma_0) &
    (\frac{8}{15}\Sigma_0) &
    (-\frac{16}{35}\Sigma_0) \\
    %%
    &&&\\
    %%
    (-\frac{2}{3}\Sigma_0) &
    (\frac{4}{9}\Sigma_0 + \frac{5}{9}\Sigma_2) &
    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
    %%
    &&&\\
    %%
    (\frac{8}{15}\Sigma_0) &
    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
    \\ 
    %%
    &&&\\
    %%
    (-\frac{16}{35}\Sigma_0) &
    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
    & 
    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
  \end{bmatrix}}\:.
  \label{eq:A_matrix}
\end{equation}
Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
the diffusive term in each equation is only for a single
'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simplified_2013},
lower order $SP_N$ approximations can be generated by setting higher
order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
yields the $SP_3$ equations). For these equations, the $SP_N$ order
defines the largest order flux moment maintained in the calculation
and the $P_N$ order defines the largest order of the cross section
expansion maintained.

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
considering the multigroup form of the $SP_N$ equations derived in
Appendix~\ref{chap:mg_spn_equations}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$, given by
\cite{evans_simplified_2013}. We again apply the change of variables
to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
To complete the spatial discretization, the gradient operators are
resolved by a finite volume scheme as in
\cite{evans_simplified_2013}. It should be noted here that with
respect to a general MCSA scheme, the introduction of fission in the
system will only affect the source vector in the linear system for the
eigenvalue schemes used in this work. The linear operator, and
therefore overall MCSA performance will be dictated by streaming and
scattering as defined on the left-hand side.

%%---------------------------------------------------------------------------%%
\section{Basic Preconditioning for the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their diffusion-like form. In this section, we will
compute the spectral radius of the iteration matrix formed by the
$SP_N$ equations while parametrically varying the $SP_N$ order, the
$P_N$ order, the number of energy groups, and
upscattering/downscattering between energy groups for Jacobi
preconditioned problems to ensure that this type of preconditioning
will be applicable to the fuel assembly problem. For each problem in
the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{c}{Parameter}& 
      \multicolumn{1}{c}{Value} \\\hline
      Mesh Element Size & 1.0 \\
      Mesh Elements & $4 \times 4 \times 4$ \\
      Reflecting Boundaries & True \\
      Materials & 1 \\
      Source Strength & 1.0 \\
      Total Cross Section & 5.0 \\
      In-Group Cross Section & 0.25 \\
      Downscatter Cross Section & 1.0 \\
      Upscatter Cross Section & 0.1 \\
      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
      parameter study.} \textit{The parameters were fixed for each
      spatial location and energy group.}}
  \label{tab:spn_fixed_parameters}
\end{table}

Mesh element sizes were the same in all cardinal directions with
reflecting conditions on each boundary. For the energy parameter, 1
and 10 energy groups were used with upscatter/downscatter varied for
the 10 group case. For the downscatter cases, all groups could scatter
to all lower energy groups with the cross sections given in
Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
groups could scatter to all other energy groups with different cross
sections for upscatter and downscatter used. A uniform isotropic
source of neutrons is given in each group as well.

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.3in]{chapters/spn_equations/group1.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
      used to show detail of the blocks formed by the discretization.}}
  \label{fig:group1}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.3in]{chapters/spn_equations/group10ds.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter only.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10ds}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.3in]{chapters/spn_equations/group10us.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10us}
\end{figure}

We note a few key features of the sparsity plots. The first is that
for multigroup problems without full upscatter and downscatter
(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
and therefore a linear solver that can handle asymmetric linear
systems is required. Nearly all problems of interest will not have
full upscattering or downscattering. Second we note the largely
diagonal character of these systems, although the blocks from
Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
$N_g\times(N+1)/2$.

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we choose
\textit{block Jacobi} preconditioning as a left preconditioner for the
system. Block Jacobi preconditioning extracts the diagonal block
elements of the matrix to use as the preconditioner as shown on the
left side of Figure~\ref{fig:block_jacobi_ex}. Shown on the right side
of Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner
is trivial with each diagonal block inverted separately.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.6}{
    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion of the
      preconditioner is trivial and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}

For high performance implementations block Jacobi preconditioning has
several attractive properties. First, the blocks in the transport
operator come from the energy/angle discretization of the transport
equation as given by Eq~(\ref{eq:A_block_matrix}). Each block on the
diagonal is bound to a mesh element in the system (note there are 8
blocks on the diagonal in each of the sparsity patterns with a mesh of
$2 \times 2 \times 2$) and therefore we expect the matrix elements
forming the block to be entirely local in domain decomposed
calculations. Second, these blocks are typically dense and nearly
lower triangular for many transport problems meaning that established
dense matrix methods can be used for fast inversion.

Spectral radius computations for the block Jacobi preconditioned
iteration matrix were performed with Table~\ref{tab:group1bj} giving
the results for the 1-group case, Table~\ref{tab:group10dsbj} for the
10-group case with full downscatter only and
Table~\ref{tab:group10usbj} for the 10-group case with full
downscatter and full upscatter. A block size of $N_g\times(N+1)/2$ was
used for the preconditioner in all cases.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1bj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dsbj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10usbj}
\end{table}

From the tabulated block Jacobi data it is clear that this is a viable
preconditioning choice for all $SP_N$ problems in terms of Monte Carlo
solution methods. All cases were observed to have a spectral radius
below unity. Based on these results, the block Jacobi method should be
the applicable to more complicated problems representative of physical
nuclear systems.

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin, control rod, and moderator materials and
the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
while a cross section of the materials configuration along the
horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. The
control rods were removed for these problems leaving only moderator in
their place. A detailed view of the assembly bottom is given in
Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
vacuum conditions are used as well as on the top and right boundaries
in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
used on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized control rods. Each fuel pin is resolved by a $2
      \times 2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is homogenized
      moderator/control rodes and darker purple/red material is
      homogenized moderator/fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is homogenized moderator/control rods and darker
      purple/red material is homogenized moderator/fuel.}}
  \label{fig:problem3_end}
\end{figure}

Significant geometric details are contained in the model including
spacer grids, fuel pins with homogenized cladding and gas gap, core
plenum, and moderator with boron. Group cross sections and other
discrete nuclear data are generated as needed by a cross-section
processing module dependent on the meshing parameters used to
discretize the geometry and single-dimension pin-cell calculations for
initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
gives the primary design parameters for the fuel assembly
calculations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Parameter}} & 
      \multicolumn{1}{l}{\textbf{Value}} \\
      Power Level & 0 MW \\
      Inlet Temperature & 326.85C \\
      Fuel Temperature & 600C \\
      Boron Concentration & 1300 ppm \\
      Moderator Density & 0.743 g/cc \\
      Helium Density & \sn{1.79}{-4} g/cc \\
      Zirconium Density & 6.56 g/cc \\
      Stainless Steel Density & 8.0 g/cc \\
      Inconel Density & 8.19 g/cc \\
      UO2 Density & 10.257 g/cc \\
      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
      assembly criticality calculation.}}
  \label{tab:problem3_parameters}
\end{table}

>>>>

\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
\label{subsec:spn_estimator_comparison}
The iterative performance of MCSA is studied in this appendix with
both the collision and expected value estimators using the ILUT
preconditioning. Although the expected value estimator achieved the
best iterative performance in analysis for the simple Poisson problem
in \S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective for
the $SP_N$ equations. The deterministic component of the expected
value estimator that couples the current state to other local states
through the iteration matrix stencil requires application of the
estimator to potentially several orders of magnitude more states at
every transition event.

For this study, the fuel assembly problem with a 1-group $SP_1$
discretization was solved using MCSA with the adjoint Monte Carlo
solver using both the collision and expected value estimators and both
the Richardson and minimal residual fixed point iterations. Using the
analysis from Appendix~\ref{chap:spn_mcsa_relaxation}, a Neumann
relaxation parameter of 0.7 was used with both estimators and fixed
point iterations. For the collision estimator, a Richardson relaxation
parameter of 1.1 was used with the Richardson iteration while it was
found the expected value estimator had the best performance when a
Richardson relaxation parameter of 1.0 was used instead. In addition,
ILUT preconditioning with a drop tolerance of \sn{1}{-5} and a fill
level of 5 was used with a reduced domain approximation fill level of
100 and a threshold of \sn{1}{-10} as in the previous
analysis. Artificial absorption was also introduced with a value of
0.2 for the MCSA calculations that used the collision estimator as CPU
timing performance was improved without a loss in iterative
performance while no absorption was used for calculations using the
expected value estimator. For each estimator, the number of stochastic
histories used to compute the MCSA correction was varied from 25 to
50,000 with the number of MCSA iterations required to converge to a
tolerance of \sn{1}{-8} recorded for a single eigenvalue iteration
along with the CPU time required to achieve convergence.
